HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-04-24 docs/HLL_exp_bad/main.tex
attack impact

\paragraph{Scenario S2:} The adversary knows $h$ so it can adjust fields in the item $x$ it is trying to insert and keep only those for which the $\ell$ rightmost bits of $h(x)$ start with a one bit. In this case, the number of leading zero bits computed by the estimator of the destination bucket is always zero, so the estimator of this bucket is unlikely to be updated and the averaged approximate cardinality is likely to be left unchanged. For example, in the scenario of \cite{portscanhll}, the adversary could change the source port of his IP packet until it is satisfied with its hash value. Given sufficient flexibility in the input $x$, the adversary can insert many items into the HLL without increasing the cardinality markedly, by repeatedly sampling from the input domain.

\paragraph{Scenario S3:} The adversary can adopt the same strategy as for S2, but make the attack more efficient and reduce the sampling requirements with the additional information available in this scenario (namely, the intermediate estimators computed per bucket). So, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increment the current bucket estimate. In other words, if $h(x)$ is mapped to the $i$-th bucket with estimated cardinality $2^{c_i + 1}$ (because a value with $c_i$ leading zeroes was previously observed for this bucket), then, instead of the strict restriction of having a one bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a one bit in any of the first $c_i+1$ positions in this substring. In case the adversary hits an empty bucket (for which the estimator is 0), it must skip that bucket completely. Under these conditions, the estimators in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.  A special case arises when the HLL is completely empty at the beginning of the attack; here the adversary is forced to increase the estimate in at least one bucket, but with sampling this can obviously be done in such a way as to increase the HLL estimate to $2$ in a single bucket, leaving all the other buckets untouched.

\subsection{Impact}
Now that we have a way to exploit the vulnerability, we need to evaluate to which extent it can affect the estimate. We assume that $h$ is a "good" hash function, so any bit of $h(x)$ has a probability of $\frac{1}{2}$ to be 1 and $\frac{1}{2}$ to be 0.\\

Under S2, the adversary need to craft 2 $x$s on average to obtain a 1 bit at the $n$th position of $h(x)$ and successfully perform the attack. As we do not care about the remaining bits, the adversary can theoretically "hide" up to 2$^{len(h(x))-1}$ items into the sketch.\\

Under S3, let's compute the average work for the adversary to find an $x$ such that $h(x)$ meets the attack's requirements. We assume $h(x)$ is mapped to bucket $i$. The probability that $h(x)$ does not increment the estimator is the probability that $h(x)$ has less than $c_i$ leading 0s. So:\\
$P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i)=P(h(x)$ has LESS than $c_i$ leading 0s$)=1-P(h(x)$ has strictly MORE than $c_i$ leading 0s$)=1-\frac{1}{2^{c_i+1}}$\\
Across all buckets, the adversary does not update any estimate with probability $P=
\sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i$ AND $h(x)$ does NOT update bucket $i) = \sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i)P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i) = \sum_{i=1}^{2^n}\frac{1}{2^n} \times (1-\frac{1}{2^{c_i+1}}) = 1-\frac{1}{2}\times(\frac{1}{2^n}\times \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) = 1-\frac{1}{2}\times\frac{1}{H_c} = 1-(2H_c)^{-1}$, with $H_c$ the harmonic mean of the numbers $2^{c_1}, 2^{c_2}, ..., 2^{c_{2^n}}$.
So the adversary needs to craft on average $(1-(2H_c)^{-1})^{-1} x$s, and can theoretically "hide" up to 2$^{len(h(x))}\times (1-(2H_c)^{-1})$ items into the sketch.

\textit{NOTE: does it take into account the special case of empty bucket ? counters are initialized to $c_i=-oo$ (or -1 in the code???) thus $P(h(x)$ has LESS than $-oo$ leading $0s) = 0$ but it is hard to add it to the probabilities while staying clear.}\\
%2$^{len(h(x))}\times \frac{2H_c-1}{H_c}$ items into the sketch.\\

We assume here that there already are values coming from other honest users' data in the sketch, we now remove this assumption. Unfortunately, the adversary in S2 has no way to detect that he is attacking an empty sketch and cannot adapt his strategy, so all buckets hit will update once their estimator to $2^{0+1}=2$ items, leading to the sketch final approximation of the number of distinct elements to $\alpha2^{n+1}$ regardless of the amount of items the adversary will send. However, the adversary in S3 is able to notice that all buckets are empty and adapt his attack. As mentioned above, the adversary in S3 avoids empty buckets, but when dealing with an empty sketch, sacrifices must be made if the adversary wants to insert at least one item. To limit the detection, the adversary can target and fill one bucket only. This would fix $n+1$ bits of $h(x)$, allowing the adversary to add $2^{len(h(x))-n-1}$ elements while keeping the cardinality to 1.

\subsection{Experimental results}
There exist many implementations of the HyperLogLog algorithm available online. Among the repositories collecting more than 350 stars on github, we can cite \cite{clahll} and \cite{datasketch}. It is also featured in many frameworks, including Redis \cite{redis}, Google's BigQuery \cite{bigquery} and several products of Apache such as Spark \cite{spahll} \cite{spahll2} and Druid \cite{druhll}. It is interesting to point out that they all use a constant number of buckets and a fixed hash function, \textit{which is public or should be public according to Kerckhoffs's principle,} thus are potentially vulnerable to cardinality estimate manipulation when put in an adversarial setting.\\

>>>>

\noindent\paragraph{Scenario S2:} The adversary knows $h$ so it can adjust fields in the item $x$ it is trying to insert and keep only those for which the $\ell$ rightmost bits of $h(x)$ start with a one bit. In this case, the number of leading zero bits computed by the estimator of the destination bucket is always zero, so the estimator of this bucket is unlikely to be updated and the averaged approximate cardinality is likely to be left unchanged. For example, in the scenario of \cite{portscanhll}, the adversary could change the source port of his IP packet until it is satisfied with its hash value. Given sufficient flexibility in the input $x$, the adversary can insert many items into the HLL without increasing the cardinality markedly, by repeatedly sampling from the input domain.

\noindent\paragraph{Scenario S3:} The adversary can adopt the same strategy as for S2, but make the attack more efficient and reduce the sampling requirements with the additional information available in this scenario (namely, the intermediate estimators computed per bucket). So, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increment the current bucket estimate. In other words, if $h(x)$ is mapped to the $i$-th bucket with estimated cardinality $2^{c_i + 1}$ (because a value with $c_i$ leading zeroes was previously observed for this bucket), then, instead of the strict restriction of having a one bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a one bit in any of the first $c_i+1$ positions in this substring. In case the adversary hits an empty bucket (for which the estimator is 0), it must skip that bucket completely. Under these conditions, the estimators in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.  A special case arises when the HLL is completely empty at the beginning of the attack; here the adversary is forced to increase the estimate in at least one bucket, but with sampling this can obviously be done in such a way as to increase the HLL estimate to $2$ in a single bucket, leaving all the other buckets untouched.

\subsection{Impact}
We now evaluate the impact of the above attacks on the HLL cardinality estimate. We assume that $h$ is a ``good" hash function, so that the bits of $h(x)$ are balanced (i.e. have a probability of $1/2$ of equalling 0 and 1).

In scenario S2, the adversary needs to craft two distinct candidates for $x$ on average to obtain a one bit at the first position in the $\ell$ rightmost bits of $h(x)$, thus successfully performing the attack for a single insertion. Assuming the adversary has $t$ bits of flexibility in its choice of inputs, the adversary can expect to ``hide''  2$^{t-1}$ distinct items without increasing the HLL cardinality estimate at all.

In scenario S3, we compute the average work required of the adversary to find an input $x$ such that $h(x)$ meets the attack's requirements. Let us assume for the moment that $c_i \ge 1$ for all buckets $i$. Assume $h(x)$ is mapped to bucket $i$. The probability that $h(x)$ does not increment the estimator is the probability that $h(x)$'s rightmost $\ell$ bits have $c_i$ or less leading zeroes. This is equal to $1-\frac{1}{2^{c_i+1}}$ assuming $h$ behaves like a random function. Averaging over all $2^n$ buckets (and assuming the bucket choice is uniformly random), the probability that input $x$ does not increment the estimator is given by:
\begin{eqnarray*}
\sum_{i=1}^{2^n}\frac{1}{2^n} \cdot (1-\frac{1}{2^{c_i+1}}) & = & 1-\frac{1}{2}\cdot (\frac{1}{2^n}\cdot \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) \\
& = &1-(2H_c)^{-1} \\
\end{eqnarray*}
where $H_c$ is the harmonic mean of the counts $2^{c_1}, 2^{c_2}, \ldots, 2^{c_{2^n}}$.

%$P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i)=P(h(x)$ has LESS than $c_i$ leading 0s$)=1-P(h(x)$ has strictly MORE than $c_i$ leading 0s$)=1-\frac{1}{2^{c_i+1}}$\\
%Across all buckets, the adversary does not update any estimate with probability $P=
%\sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i$ AND $h(x)$ does NOT update bucket $i) = \sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i)P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i) = \sum_{i=1}^{2^n}\frac{1}{2^n} \times (1-\frac{1}{2^{c_i+1}}) = 1-\frac{1}{2}\times(\frac{1}{2^n}\times \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) = 1-\frac{1}{2}\times\frac{1}{H_c} = 1-(2H_c)^{-1}$, with $H_c$ the harmonic mean of the numbers $2^{c_1}, 2^{c_2}, ..., 2^{c_{2^n}}$.
So the adversary needs to craft on average $(1-(2H_c)^{-1})^{-1}$ distinct candidates for $x$ in order to insert a single one, and, given $t$ bits of flexibility in its choice of inputs, can expect to ``hide" up to $(1-(2H_c)^{-1}) \cdot 2^t$ distinct items without increasing the HLL cardinality estimate at all.

%\textit{NOTE: does it take into account the special case of empty bucket ? counters are initialized to $c_i=-oo$ (or -1 in the code???) thus $P(h(x)$ has LESS than $-oo$ leading $0s) = 0$ but it is hard to add it to the probabilities while staying clear.}\\
%2$^{len(h(x))}\times \frac{2H_c-1}{H_c}$ items into the sketch.\\

Note that the above analysis made the assumption that every bucket count $c_i$ was already at least 1. When some bucket counts are zero, the above analysis needs to be modified in order to avoid touching those buckets altogether. The modified analysis is quite straightforward. Suppose $r \le 2^n-1$ out of $2^n$ buckets have a count of zero. Then the cost of crafting an input $x$ becomes $(1-(2H_c)^{-1})^{-1} \cdot (1- \frac{r}{2^n})^{-1}$ trials, while the total number of items the adversary can expect to insert becomes $(1-(2H_c)^{-1}) \cdot (1- \frac{r}{2^n})\cdot 2^t$.

We assumed so far that there are already values coming from other honest users' data in the HLL. We now remove this assumption. Unfortunately, the adversary in scenario S2 has no way to detect that he is attacking an empty HLL and cannot adapt his strategy, so all buckets hit will update their estimator once to $2^{0+1}=2$ items, leading to a final cardinality approximation of $\alpha2^{n+1}$ regardless of the number of items the adversary inserts. However, the adversary in S3 is able to detect that all buckets are empty and adapt his attack accordingly. The adversary can target and fill one bucket only. This requires fixing $n+1$ bits of $h(x)$ for each input $x$, allowing the adversary to ``hide'' $2^{t-n-1}$ elements while keeping the HLL cardinality estimate to 1. Other work/cardinality estimate trade-offs are of course possible.

\subsection{Experimental Results}
There exist many implementations of the HyperLogLog algorithm available online. Among the repositories collecting more than 350 stars on github, we can cite \cite{clahll} and \cite{datasketch}. It is also featured in many frameworks, including Redis \cite{redis}, Google's BigQuery \cite{bigquery} and several products of Apache such as Spark \cite{spahll} \cite{spahll2} and Druid \cite{druhll}. It is interesting to point out that they all use a constant number of buckets and a fixed hash function, \textit{which is public or should be public according to Kerckhoffs's principle,} thus are potentially vulnerable to cardinality estimate manipulation when put in an adversarial setting.\\
